\documentclass[12pt,noauthor,nooutcomes]{ximera}%% SEE: https://www.overleaf.com/project/626c378b4e3e27903576f685
\author{}

\input{../preamble.tex}



\title{Project ideas}

\begin{document}
\begin{abstract}
    Some ideas for projects.
\end{abstract}
\maketitle

\begin{project}
Explore fractals, both in mathematics and in music.

Reference: \url{https://youtu.be/VSZTEIMlVmY}

Another example: \textit{Structural scaling in Bach's Cello Suite \#3} \url{http://mathtourist.blogspot.com/2008/09/fractal-in-bachs-cello-suite.html}
\end{project}

\begin{project}
Explore the work of Iannis Xenakis, his stochastic music and his game theory piece \textit{Duel y Strategie}.
\end{project}

\begin{project}
For students who have/play an instrument: Explore the sound waves produced by that instrument, observe the decomposition in harmonics.

Reference: \url{https://youtu.be/ybPvR0mraaA}

We would need to find a good analyzer online that can do this. Or code it into python.
%See https://publish.illinois.edu/augmentedlistening/tutorials/music-processing/tutorial-1-introduction-to-audio-processing-in-python/   and   https://www.openbookproject.net/py4fun/wave/wave.html

\end{project}

\begin{project}
Build another instrument and explore the different notes it produces (both from a musical and a theoretical point of view).
\end{project}

\begin{project}
Discuss various tuning methods, ``just intonation,''  ``Pythagorean tuning,'' and ``equal temperament.''

References include the AMS book and the video
\url{https://youtu.be/rdIxmRLPS_M}
\end{project}

\begin{project}
Composition with magic squares from the \textit{From Music to Mathematics} book.

\url{http://www.frommusictomath.com/uploads/4/3/3/2/43320843/daviesmagicsquares.pdf}
\end{project}

\begin{project}
Explain why guitars sometimes use ``fan frets.''
\end{project}


\begin{project}
Make your own math-inspired composition!
\end{project}


\begin{project}
For students who like coding: do something on SonicPy.  (Or Pico-8)
\end{project}


\begin{project}
The OSUMB marching drills have mathematical content one could discuss.
\end{project}


\begin{project}
Approximating real numbers with, continued fractions explaining why there are 12 tones in the western scale. 
\end{project}


\begin{project}
Discuss Shepard's Tones and Risset Rhythms. 
\end{project}


\begin{project}
Why is a trumpet in $B\flat$?
\end{project}


\begin{project}
Create hearing test not dependent on pure tones.
\end{project}


\begin{project}
An analysis of Kim Harmonics. See \url{https://www.youtube.com/watch?v=gQhQ06kXt0}
\end{project}


\begin{project}
Scales for classical stringed instruments, viewed as covering spaces.
\end{project}


\begin{project}
Give a discussion of nonwestern scales.
\end{project}


\begin{project}
Zipf's law and music: Take a largish piece and count the frequency of notes. 
\end{project}


\begin{project}
Explain the Janko Piano.
\end{project}



\begin{project}
Ciphers and music, think about a Caesar cipher and music. 

\url{https://www.wired.com/story/merryl-goldberg-music-encryption-ussr-phantom-orchestra/}
\end{project}

\begin{project}
How do touch tone phones work?
\end{project}


\begin{project}
D.J.\ and subwoofer location. 
\end{project}




\begin{project}
Copyright and music, see:
\begin{enumerate}
\item \url{https://youtu.be/HkUgpfZ3rjQ}
\item \url{https://youtu.be/1wvhIF5tJ9o}
\item \url{https://youtu.be/axRMZqUNVEw}
\end{enumerate}
\end{project}


\begin{project}
Tartini tones, what are they? 
\end{project}


\begin{project}
Fanned frets.
\end{project}




\end{document}