\documentclass[12pt,noauthor,nooutcomes,handout,hints]{ximera}

\input{../preamble.tex}

\author{}


\title{Hertz and cents}

\begin{document}
\begin{abstract}
  We learn about ways of measuring pitch. At the same time, we introduce the idea of octaves.
\end{abstract}
\maketitle

\begin{instructorNotes}
This activity is about measurement. 
\begin{itemize}
\item Notes measured by frequency are on an exponential scale. 
\item Notes measured by cents are on a linear scale. 
\item Frequency is an absolute scale.
\item Cents is a relative scale.
\item Octaves are notes whose frequencies have a ratio of $2^n$ ($n\in\Z$) and whose cents difference are a multiple of $1200$).

\end{itemize}


%% see https://www.youtube.com/watch?v=B3QWCrRjkb0
%% see https://pages.mtu.edu/~suits/notefreqs.html


\end{instructorNotes}


\begin{question}
A piano typically has $88$ keys. Here we show the notes corresponding to some of these keys (keys $37$--$61$) along with their frequency in Hertz. Remember that Hertz are the same as cycles or full waves per second.
\noStaff



As you would expect from what we learned before, each note has a different frequency.  Note that this diagram includes more notes than what we had before. Some of the notes are named with the same letters, but they have a numerical subindex to distinguish one from the other.

\begin{enumerate}
%\item What's the \textbf{difference} in frequency between $A_4$ and $A_3$?

%\answerspace

%\item What's the \textbf{difference} in frequency between $A_5$ and $A_4$?


%\answerspace


\item What's the \textbf{ratio} of the frequencies between $A_4$ and $A_3$?

\answerspace


\item What's the \textbf{ratio} of the frequencies between $A_5$ and $A_4$?

\answerspace

\item What's the \textbf{ratio} of the frequencies between $E_5$ and $E_4$?

\answerspace

\item Choose any other two octave notes (two notes with the same letter name but different numbers) and compute the ratio of their frequencies.

\answerspace

\item What pattern do you notice?

\answerspace

\end{enumerate}
\end{question}

\begin{question} Look at your work above.
\begin{enumerate}

\item Write the sequence of frequencies (in Hz) of the notes $A_i$ where $i=1,2,3,4,5,6,7$ (that is $A_1, A_2, A_3$, etc.).

\answerspace


    \item What do you think would be the frequency in Hertz of the notes $E_3$, $B_5$, and $C_6$? \textbf{Explain your reasoning.}

\answerspace


%\item Remember what you've learned at school. Do you know what a sequence like this, that grows by a factor of $2$, is called?

%\answerlines

\begin{instructorNotes}
In grade school this would be called a geometric sequence, but some students might not remember or even have learned that. If they don't recall this, they can simply skip the question.

The goal is not to know a technical term, but to make a connection with something familiar.
\end{instructorNotes}

%\item The distance between one note and the next note with the same letter is called an octave. For example, we might say that $A_5$ is one octave above $A_4$ and two octaves above $A_3$. 

%Our ears (and brains) perceive doubling or halving frequencies as the same note. That is the reason why the notes with frequencies $55$ Hz, $110$ Hz, $220$ Hz, $440$ Hz, $880$ Hz are all called $A$s (and same goes for the other letters). 

%Look at the piano diagram, why do you think this is called an octave? \begin{hint}
%The word \textit{octave} means eighth.\end{hint}

%\answerlines
\end{enumerate}
\end{question}


\begin{question}
Another way of assigning numbers to pitches is to use the idea of a ``cent.'' We define the difference between any two adjacent piano keys (including white and black keys) to be $100$ cents. 
\begin{enumerate}
\item What's the \textbf{difference} in cents between $A_4$ and $A_3$?

\answerlines

\item What's the \textbf{difference} in cents between $A_5$ and $A_4$?


\answerlines


\item What's the \textbf{ratio}
\[
\frac{\mathrm{cents}(A_5-A_3)}{\mathrm{cents}(A_4-A_3)}?
\]

\answerlines


\item What's the \textbf{ratio}
\[
\frac{\mathrm{cents}(G_5-A_3)}{\mathrm{cents}(G_4-A_3)}?
\]


\answerlines

\end{enumerate}


\begin{question}
Compare your work in questions 2 (frequencies) and 4 (cents).
\begin{enumerate}
    \item When was the difference preserved?
    \item When was the ratio preserved?
\end{enumerate}
\end{question}


%While frequency is an \textit{absolute scale}, cents are a \textbf{relative scale}, meaning it only makes sense (pun intended!) to think of the \textbf{difference}, or \textbf{contrast}, in the cents of a pitch. 
Frequency is an \textbf{absolute scale}, meaning each note gets a unique number of Hertz for its frequency, different than the frequency of any other note. In contrast, cents make a \textbf{relative scale}, since they refer to \textbf{differences}, or contrast, between the notes. The same number of cents repeats between several pairs of notes. For example, you can see a 200 difference in many different places on the piano. 

\end{question}

\begin{question} %% students will need to figure out A_2
Now we are going to make \textbf{two plots}, one with frequency and the other with cents.
\begin{enumerate}
    \item Plot the frequency of each note.
    
\begin{tikzpicture}
\begin{axis}[
x=1.5cm,y=0.005cm,
axis lines=middle,
ymajorgrids=true,
xmajorgrids=true,
xmin=0,
xmax=8,
ymin=0,
ymax=3600,
xtick={-0.0,1.0,...,8.0},
%xticklabels={0,$A_1$,,$B_3$,$C_4$,,$D_4$,,$E_4$,$F_4$,,$G_4$,,$A_4$,,$B_4$,$C_5$,,$D_5$,,$E_5$,$F_5$,,$G_5$,,$A_5$},
xticklabels={0,$A_1$,$A_2$, $A_3$, $A_4$, $A_5$, $A_6$, $A_7$},
ytick={200.0,400.0,...,3600.0},]
\end{axis}
\end{tikzpicture}

\newpage
\item Now plot the cents difference comparing against the first note on our piano above. For each note, plot the cents difference between that note and the $A_3$ note.

\begin{tikzpicture}
\begin{axis}[
x=1.5cm,y=0.0025cm,
axis lines=middle,
ymajorgrids=true,
xmajorgrids=true,
xmin=0,%-0.7727900000000003,
xmax=8,
ymin=0,%-99,
ymax=7500,
xtick={-0.0,1.0,...,8.0},
%xticklabels={0,$A_3$,,$B_3$,$C_4$,,$D_4$,,$E_4$,$F_4$,,$G_4$,,$A_4$,,$B_4$,$C_5$,,$D_5$,,$E_5$,$F_5$,,$G_5$,,$A_5$},
xticklabels={0,$A_1$,$A_2$, $A_3$, $A_4$, $A_5$, $A_6$, $A_7$},
ytick={0.0,600.0,...,7200.0},]
\end{axis}
\end{tikzpicture}
\newpage
\item Describe the plots in your own words. What do they mean? How do they relate to each other?

\answerlines

\answerlines
\item What is the cents difference between octave notes?

\answerlines

\item How does the frequency change between octave notes?

\answerlines

\item How do you think the answer to the previous two question relates to the plots?

\answerlines

\end{enumerate}
\end{question}



\end{document}
