\documentclass[12pt]{ximera}

\input{../preamble.tex}
\title{References and further reading}

\begin{document}
\begin{abstract}
  These are the sources we used to guide us in the construction of this
  document.
\end{abstract}
\maketitle

%\vspace{-1in}

\begin{thebibliography}{MMMM2002}


\bibitem[FFW2010]{FFW2010} Fauvel, Flood, Wilson (Editors).
\textbf{Music and mathematics.
From Pythagoras to fractals.}  Oxford University Press, Oxford, England 2010. x+189 pp.\begin{annotate}
    A collection of separate articles that provide an excellent first introduction to mathematics and music.
  \end{annotate}



\bibitem[rG2016]{rG2016} Roberts.
\textbf{From music to mathematics.
Exploring the connections.}  Johns Hopkins University Press, Baltimore, MD, 2016. xviii+301 pp.
  \begin{annotate}
    A undergraduate mathematics textbook, that takes a  ``music first'' approach, very complete in its mathematical analysis.
  \end{annotate}



\bibitem[bH2009]{bH2009} Hopkin.
\textbf{Making musical instruments with kids. 67 easy projects for adults working with children.}  See Sharp Press, Tucson, AZ, 2009. iv+116 pp.
  \begin{annotate}
    Thoughtful designs for fun instruments that can be built at home.
  \end{annotate}



\bibitem[MMP2016]{MMP2016} Mazzola, Mannone, Pang.
\textbf{Cool math for hot music.
A first introduction to mathematics for music theorists.} Computational Music Science. Springer, 2016. xv+323 pp.
  \begin{annotate}
    Advanced ideas in mathematics illustrated though example from music.
  \end{annotate}
  
\bibitem[dW2009]{dW2009} Wright. \textbf{Mathematics and music.}
Mathematical World, 28. American Mathematical Society, Providence, RI, 2009. xiv+161 pp.
  \begin{annotate}
    A foundational book in mathematics in music, the key mathematical ideas are clearly explained at an undergraduate level. 
  \end{annotate}
  
  

  
\end{thebibliography}
\end{document}