\documentclass[12pt,noauthor,nooutcomes]{ximera}

\author{}

\input{../preamble.tex}



\title{Tonic, dominant, octave}

\begin{document}
\begin{abstract}
    We learn the tonic, dominant, octave.
\end{abstract}
\maketitle
\begin{instructorNotes}
LEARNING OUTCOME: OCTAVES AND PERFECT FIFTHS SOUND BEST.


\end{instructorNotes}


Throughout history, a relationship between three pitches has captivated humans more than any other. For example, they are the first three notes of this piece

\textit{Also Sprach Zarathustra}, by Strauss
\begin{center}
\url{https://youtu.be/dfe8tCcHnKY}
\end{center}
Each pitch in this relation has a special name. The lowest is called the \textbf{tonic}, and is the first pitch your heard above. Next is the \textbf{dominant}, the middle pitch. Finally, the highest pitch is the \textbf{octave}. These different pitches are intimately connected to overtones. 

Recall that the most basic harmonics are the 1st harmonic, 2nd harmonic, and 3rd harmonic and they look like this:

\begin{center}
\begin{tabular}{ccc}
\begin{tikzpicture}
  \wave{1}
\end{tikzpicture} 
& \begin{tikzpicture}
  \wave{2}
\end{tikzpicture} &
\begin{tikzpicture}
  \wave{3}
\end{tikzpicture} \\
$1$st harmonic & $2$nd harmonic & $3$rd harmonic \\ 
\end{tabular}
\end{center}
\begin{itemize}
\item The frequency of the tonic is the frequency of the $1$st harmonic.
\item The frequency of the dominant is the frequency of the $3$rd harmonic.
\item The frequency of the octave is the $2$nd harmonic.
\end{itemize}
The trio tonic - dominant - octave is a relation between notes, not a specific set of notes. What you choose as your tonic, will determine what the dominant and the octave are.


\begin{question}
Let's remind ourselves, if the $1$st harmonic has frequency $F$, what are the frequencies of the $2$nd and $3$rd harmonics?

Here's our piano diagram again.
\noStaff
Let's take $C_4$ to be our tonic.
\begin{enumerate}
\item In this case, what is the frequency of the tonic?
\item What frequency corresponds to the dominant? What notes is this closest to?  
\item What frequency corresponds to the octave? What notes is this closest to?  
\end{enumerate}
\end{question}



\begin{question}-
\begin{enumerate}
\item Suppose we have a string of length $L$ and we make a second string whose length is $2/3$ that of the original. What would the frequency of this new string's $1$st harmonic be?
\item A string of length $L$ sounds pleasant when plucked along with a string of length $\frac{L}{2}$. Explain why this is in terms of overtones.
\item A string of length $L$ sounds pleasant when plucked along with a string of length $2\frac{L}{3}$. Explain why this is in terms of overtones.
\end{enumerate}
\end{question}





\begin{question}
Your friend \textit{Mathemusical Mac} suddenly has a brilliant idea:
\begin{quote}
``Take your favorite note on the piano, take it as the tonic, and play the dominant above it. Repeating this process will hit everynote on the piano.''
\end{quote}
\begin{enumerate}
    \item What's \textit{Mathemuscial Mac's} reasoning behind this claim?

\item The power of math is that nobody needs to tell you what's right and what's wrong, you just compute and figure it out. Is \textit{Mathemusical Mac's} claim correct? Compute and figure it out!
\end{enumerate}
\end{question}









\end{document}