\documentclass[12pt,noauthor,nooutcomes,handout]{ximera}

\author{}

\input{../preamble.tex}



\title{Enharmonic equivalents}

\begin{document}
\begin{abstract}
  We introduce enharmonic equivalents.
\end{abstract}
\maketitle





Let's see our section of piano again:

\staffBack



%As a general resource use

%\url{https://www.youtube.com/watch?v=5aDP7u78EIU}

\begin{question}
As we saw before, a \textbf{sharp} ($\sharp$) will raise a pitch by $100$ cents. We can also use \textbf{flats} ($\flat$), which will lower a pitch by $100$ cents. For example, $D_4\flat$ is 100 cents lower than $D_4$. Hence, $D_4\flat$ is the black key to the left of the $D_4$.  

Compute how many cents are between the following pairs of notes.
\begin{enumerate}
\item $A_4$ and $A_5\flat$
\item $A_5$ and $A_5\sharp$
\item $E_4\flat$ and $F_4\sharp$
\item $D_4\sharp$ and $E_4\flat$
\item $E_4\sharp$ and $F_4$
\end{enumerate}
\end{question}



\begin{question}
   Notes are called \textbf{enharmonic} when they correspond to the same piano key, but are ``spelled'' differently.
   \begin{enumerate}
   \item What's another spelling for $E_4\sharp$?
   \item What's another spelling for $G_4\flat$?
   \item Remember that each \# increases the note by 100 cents. What key on the piano would correspond to the note $A_4\sharp\sharp$?
   \item Remember that each $\flat$ decreases the note by 100 cents. What key on the piano above would correspond to the note $E_4\flat\flat\flat$?
   %\item Give three different spellings for $C_5$.
   
   %\answerspace
   
   
   \item Give three different spellings for $E_4$.
   
      \answerspace

   
   \end{enumerate}
\end{question}


\begin{instructorNotes}
The following questions will most likely be dicussed as a class, rather than worked individually. 

Most middle schoolers are likely not familiar with algebra yet. However, if a student finishes the previous part before the rest of the class, they can continue to these two questions, specially if they are familiar with algebra.
\end{instructorNotes}

\begin{question}
A lot of people are uncomfortable with fractions. One of the reasons I think people are sometimes uncomfortable with fractions is because there are multiple ways of writing, or ``names,'' for the same value
\[
\frac{2}{3} = \frac{4}{6} = \frac{10}{15}
\]
These fractions are equivalent, but someone might call them ``enharmonic fractions.''  Perhaps for that reason, many people like to use decimals more than fractions. However, did you know, there are ``enharmonic'' decimals too? 
\begin{enumerate}
%\item From what we explained about enharmonic fractions, fractions that represent the same value even if they are written in a different way, what would you think ``enharmonic decimals'' might be?

%\answerspace




\item Suppose that $x= 0.99999\dots$, a non-terminating decimal. Compute $y = 10x$.

\answerspace



\item Now compute $y - x =$

\answerspace

\item As you know $10x - x = 9x$. So
\[
    9x = 10x - x = y - x =
\]
Plug your result from the previous question here.

\answerspace

\item  Explain why $9x = 9$, and solve for $x$.


\answerspace


\item What is the value of $x$ then? Write a conclusion about the previous questions.


\answerspace



\end{enumerate}
\end{question}

\begin{instructorNotes}
Before letting students move to the next question, make sure they conclude in the previous question that $1 = 0.999\dots$, since they are both $x$.

\end{instructorNotes}

\begin{question} From what you wrote before, find an alternative way of writing the decimal $0.1$ as an infinite decimal.

That number and $0.1$ would be ``enharmonic decimals.''
\begin{enumerate}
\item Write down your favorite finite decimal number and an enharmonic decimal with infinite digits.

\answerspace

\item Explain how every finite decimal has an infinite enharmonic decimal of the same value. 

\answerspace

\answerspace



\end{enumerate}
    
\end{question}






%% extra question
%\begin{question}
%Consider the following excerpt from Rossini's \textit{The Barber of %Seville Overture}:

%\begin{music}
%\instrumentnumber{1}            % number of instruments
%\setstaffs1{1}              % number of lines per instrument
%\generalmeter{\meterfrac{4}{4}}   % time stamp. meterC is 4/4, allabreve is 2/2 or cut time.
%\generalsignature{4}            % sets the key. a number greater than 0 is sharp, smaller than 0 is flat
%\startextract
% bar 1
%  \Notes \qa{>f} \qa{g} \qa{^f} \qa{g} \en 
%\endextract
%\end{music}
%The funny symbol before the first $F$ is a \textit{double sharp}. A double sharp raises the pitch $200$ cents. Speculate on why someone might write this using a double sharp.
%\end{question}






\end{document}